---
title: "Getting Started with omie2fdata"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with omie2fdata}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

# Overview

`omie2fdata` follows a three-layer workflow:

1. raw acquisition (`omie2txt()`, `omie2download()`),
2. a common tabular representation (`omie2df()`),
3. functional conversion (`omie2fdata()`, `omie2ldata()`).

This design means that the same parser is used whether the raw OMIE file was
saved previously or downloaded temporarily inside the conversion call.

# Consolidated market-results source

The original package workflow is retained as `source = "results"`.

```{r}
fd <- omie2fdata(
  start_date = "2025-04-26",
  end_date = "2025-04-30",
  var_names = "price_es",
  resolution = "hour"
)
plot(fd)
```

For several variables:

```{r}
ld <- omie2ldata(
  start_date = "2025-04-26",
  end_date = "2025-04-30",
  var_names = c("price_es", "buy_es", "sell_es")
)
plot(ld$price_es)
plot(ld$buy_es)
```

# Reusing previously downloaded files

Raw files do not need to be downloaded again. A file, a vector of files, or a
directory can be supplied directly. This is useful for archives created with an
earlier version of `omie2fdata`.

The package also contains a few small raw OMIE examples for tests and offline
examples:

```{r}
f <- system.file("extdata", "omie_2025-04-01.txt", package = "omie2fdata")

ld_local <- omie2ldata(
  start_date = "2025-04-01",
  end_date = "2025-04-01",
  var_names = c("price_es", "buy_es"),
  resolution = "hour",
  file = f
)

plot(ld_local$price_es)
plot(ld_local$buy_es)
```

# Direct repository source

`omie2download()` downloads daily files from OMIE's public **Acceso a
ficheros** repository. The first direct source integrated with functional
conversion is `marginalpdbc`.

```{r}
paths <- omie2download(
  file_type = "marginalpdbc",
  start_date = "2026-07-01",
  end_date = "2026-07-07",
  output_dir = "data/marginalpdbc"
)
```

These local files can then be converted directly:

```{r}
fd <- omie2fdata(
  start_date = "2026-07-01",
  end_date = "2026-07-07",
  var_names = "price_es",
  resolution = "15min",
  source = "marginalpdbc",
  file = "data/marginalpdbc"
)
plot(fd)
```

The package also bundles one unchanged `marginalpdbc` file, so the direct
repository parser can be demonstrated without network access:

```{r}
f_marg <- system.file(
  "extdata", "marginalpdbc_20260727.1", package = "omie2fdata"
)

ld_marg <- omie2ldata(
  start_date = "2026-07-27",
  end_date = "2026-07-27",
  var_names = c("price_es", "price_pt"),
  resolution = "15min",
  source = "marginalpdbc",
  file = f_marg
)

plot(ld_marg$price_es)
plot(ld_marg$price_pt)
```

Files can also be downloaded temporarily in the same conversion call by
leaving `file = NULL`.

# Common `data.frame`

`omie2df()` exposes the normalized intermediate representation. It contains one
row per day and market period with `date`, `period`, `argval` and the requested
variables. Daily metadata are stored in `attr(x, "metadata")`.

```{r}
x <- omie2df(
  start_date = "2026-07-27",
  end_date = "2026-07-27",
  var_names = c("price_es", "price_pt"),
  resolution = "15min",
  source = "marginalpdbc"
)
head(x)
attr(x, "metadata")
```

# A full year as functional data

Daily price and market-energy profiles are a natural FDA representation. A
closely related OMIE example was used by Febrero-Bande, González-Manteiga and
Oviedo de la Fuente (2019), who analysed daily hourly profiles of market price
and energy/demand in the Iberian electricity market.

For example, a full year of hourly OMIE results can be downloaded once and
reused locally:

```{r}
raw_2025 <- "data/omie_2025_hour"

omie2txt(
  start_date = "2025-01-01",
  end_date = "2025-12-31",
  resolution = "hour",
  output_dir = raw_2025
)

omie2025 <- omie2ldata(
  start_date = "2025-01-01",
  end_date = "2025-12-31",
  var_names = c("price_es", "buy_es"),
  resolution = "hour",
  input_dir = raw_2025
)

plot(omie2025$price_es)
plot(omie2025$buy_es)
```

This produces one daily curve per market day. `buy_es` is OMIE's total purchase
series for the Spanish system; it should not be labelled as physical system
load or demand without qualification.

The 28 April 2025 blackout can then be inspected directly from the same annual
object, for example by locating the corresponding row in `omie2025$df` and
plotting the selected daily curves.

# Resolution and daylight saving time

The default is hourly. Quarter-hourly data are accepted only from 2025-10-01.
Daily curves are normalized to 24 or 96 points. Metadata retain
`n_periods_raw`, `dst_adjusted`, `dst_type` and `aggregation` so this
normalization is explicit.

# Historical prices

Legacy files with `Precio marginal (Cent/kWh)` are recognized and converted to
`EUR/MWh`, allowing old OMEL/OMIE price curves to use the same unit as current
prices.

# OMIE data source and citation

OMIE states in its legal notice that public and free information on its website
may be used freely provided that its original content is respected, and that
OMIE must be cited as the source when the information is used in its original
format. The small files distributed in `inst/extdata` are therefore kept as raw
examples, with provenance documented in `inst/extdata/README.md`.

A suitable source citation is, for example:

> OMI-Polo Español, S.A. (OMIE). *OMIE electricity-market public data*,
> market-results or file-access dataset, https://www.omie.es/, accessed
> [date].

For a publication, cite both the `omie2fdata` package and the underlying OMIE
data source. The package code is distributed under GPL-2; this does not replace
OMIE's terms for the source market data.

# Reference

Febrero-Bande, M., González-Manteiga, W. and Oviedo de la Fuente, M. (2019).
Variable selection in functional additive regression models. *Computational
Statistics*, 34, 469-487. doi:10.1007/s00180-018-0844-5.
